\singlespacing A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}

### 2020-07-09 article.Rmd
Made some minor edits

To show this, consider the following matrix equation of the variances and the nonredundant covariances in a three-wave intercept-only model that follows directly from Equation \eqref{eq:wide} (assuming $\Cov(\varepsilon_{t1},\varepsilon_{si}) = 0, \ t \ne s$), and which we can solve easily with least squares:
\begin{align}
\bm{b} & = \bm{A}\bm{x} \\
\begin{pmatrix}
\Var(y_{1}) \\
\Cov(y_{2},y_{1}) \\
\Cov(y_{3},y_{1}) \\
\Var(y_{2}) \\
\Cov(y_{3},y_{2}) \\
\Var(y_{3})
\end{pmatrix} & = 
\begin{blockarray}{ccccc}
 & \psi & \phi_{1} & \phi_{2} & \phi_{3}  \\
 \begin{block}{c(cccc)}
 y_{1}^{2}  & 1 & 1 & 0 & 0  \\
 y_{2}y_{1} & 1 & 0 & 0 & 0  \\
 y_{3}y_{1} & 1 & 0 & 0 & 0  \\ 
 y_{2}^{2}  & 1 & 0 & 1 & 0  \\
 y_{3}y_{2} & 1 & 0 & 0 & 0  \\
 y_{3}^{2}  & 1 & 0 & 0 & 1  \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
\psi \\
\phi_{1} \\
\phi_{2} \\
\phi_{3}
\end{pmatrix}
\end{align}
where $\psi = \Var(\alpha)$, $\phi_{t} = \Var(\varepsilon_{t})$. We can solve this equation to show
\begin{align}
\bm{b} & = \bm{A}\bm{x} \\
(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{A}\bm{x} \\
(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} & = \bm{x} \\
\begin{pmatrix}
.33\Cov(y_{2},y_{1}) + .33\Cov(y_{3},y_{1}) + .33\Cov(y_{3},y_{2}) \\
\Var(y_{1}) - \psi \\
\Var(y_{2}) - \psi \\
\Var(y_{3}) - \psi
\end{pmatrix} & = 
\begin{pmatrix}
\psi \\
\phi_{1} \\
\phi_{2} \\
\phi_{3}
\end{pmatrix}.
\end{align}
This shows that if our assumption of constant covariances between the dependent variable across time holds, the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time. Once we add in observed covariates, the estimated covariance of $\alpha$ then become the \textit{conditional} covariance of $y$ over time, given those covariates.

@@

\begin{align}
\begin{split}
\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} & = 
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
 y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
 y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
 x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} + 
\begin{pmatrix}
\varepsilon_{1} \\
\varepsilon_{2} \\
\varepsilon_{3} \\
x_{1} = \delta_{1} \\
x_{2} = \delta_{2} \\
x_{3} = \delta_{3} \\
\alpha = \xi \\
\end{pmatrix}. 
\end{split} \label{eq:eta}
\end{align} 
Notice in $\bm{\zeta}^{+}$, for the independent variables we could either write, for example $x_{1}$ or $\delta_{1}$. As mentioned above, this is due to the model notation treating the independent variables like dependent variables with variances/covariances to be estimated. For the sake of simplicity, we will ignore this subtlety and refer to the observed variable from now on, keeping in mind that if the multivariate normality assumption holds, the estimated statistics will likely be sufficiently close to the sample ones for it to not make much of a difference.

@@

The key to a FE model, as opposed to an RE model are our assumptions about the relatedness of our covariate and the individual effects, i.e., $\E[x_{t}\alpha]$. For a FE model, we want to partial out any potential covariance between the independent variable and the individual effects. This accounts for any linear relationship between $x_{t}$ and the unit-specific characteristics influencing the dependent variable. Further, allowing unrestricted covariances between the independent variable itself over time will not affect how the coefficient $\beta$ is estimated, but will have an effect on the standard errors. To mimic the behaviour of a conventional FE model, we allow the independent variable to be correlated with the individual effects and itself over time. Covariances (including covariances between a variable and itself, i.e., variances) are specified using the `~~` operator:

>>>>

To show this, consider the following matrix equation of the variances and the nonredundant covariances in a three-wave intercept-only model that follows directly from Equation \eqref{eq:wide} (assuming $\Cov(\varepsilon_{ti},\varepsilon_{si}) = 0, \ t \ne s$), and which we can solve easily with least squares:
\begin{align}
\bm{b} & = \bm{A}\bm{x} \\
\begin{pmatrix}
\Var(y_{1}) \\
\Cov(y_{2},y_{1}) \\
\Cov(y_{3},y_{1}) \\
\Var(y_{2}) \\
\Cov(y_{3},y_{2}) \\
\Var(y_{3})
\end{pmatrix} & = 
\begin{blockarray}{ccccc}
 & \psi & \phi_{1} & \phi_{2} & \phi_{3}  \\
 \begin{block}{c(cccc)}
 \Var(y_{1})  & 1 & 1 & 0 & 0  \\
 \Cov(y_{2},y_{1}) & 1 & 0 & 0 & 0  \\
 \Cov(y_{3},y_{1}) & 1 & 0 & 0 & 0  \\ 
 \Var(y_{2})  & 1 & 0 & 1 & 0  \\
 \Cov(y_{3},y_{2}) & 1 & 0 & 0 & 0  \\
 \Var(y_{3})  & 1 & 0 & 0 & 1  \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
\psi \\
\phi_{1} \\
\phi_{2} \\
\phi_{3}
\end{pmatrix}
\end{align}
where $\psi = \Var(\alpha)$, $\phi_{t} = \Var(\varepsilon_{t})$. We can solve this equation to show
\begin{align}
\bm{b} & = \bm{A}\bm{x} \\
(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} & = (\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{A}\bm{x} \\
(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}\bm{b} & = \bm{x} \\
\begin{pmatrix}
.33\Cov(y_{2},y_{1}) + .33\Cov(y_{3},y_{1}) + .33\Cov(y_{3},y_{2}) \\
\Var(y_{1}) - \psi \\
\Var(y_{2}) - \psi \\
\Var(y_{3}) - \psi
\end{pmatrix} & = 
\begin{pmatrix}
\psi \\
\phi_{1} \\
\phi_{2} \\
\phi_{3}
\end{pmatrix}.
\end{align}
This shows that if our assumption of constant covariances between the dependent variable across time holds, the estimated variance of $\alpha$ is just what it should be: the average covariance between units of $y$ over time. Once we add in observed covariates, the estimated covariance of $\alpha$ then become the \textit{conditional} covariance of $y$ over time, given those covariates.

@@

\begin{align}
\begin{split}
\bm{\eta}^{+} & = \bm{B}\bm{\eta}^{+} + \bm{\zeta}^{+} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} & = 
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
 y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
 y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
 x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} + 
\begin{pmatrix}
\varepsilon_{1} \\
\varepsilon_{2} \\
\varepsilon_{3} \\
x_{1} = \delta_{1} \\
x_{2} = \delta_{2} \\
x_{3} = \delta_{3} \\
\alpha = \xi \\
\end{pmatrix}. 
\end{split} \label{eq:eta}
\end{align} 
Notice in $\bm{\zeta}^{+}$, for the independent variables we could either write, for example $x_{t}$ or $\delta_{t}$. As mentioned above, this is due to the model notation treating the independent variables like dependent variables with variances/covariances to be estimated. For the sake of simplicity, we will ignore this subtlety and refer to the observed variable from now on, keeping in mind that if the multivariate normality assumption holds, the estimated statistics will likely be sufficiently close to the sample ones for it to not make much of a difference.

@@

The key to a FE model, as opposed to an RE model are our assumptions about the relatedness of our covariate and the individual effects, i.e., $\E[x_{t}\alpha]$. For an FE model, we want to partial out any potential covariance between the independent variable and the individual effects. This accounts for any linear relationship between $x_{t}$ and the unit-specific characteristics influencing the dependent variable. Further, allowing unrestricted covariances between the independent variable itself over time will not affect how the coefficient $\beta$ is estimated, but will have an effect on the standard errors. To mimic the behaviour of a conventional FE model, we allow the independent variable to be correlated with the individual effects and itself over time. Covariances (including covariances between a variable and itself, i.e., variances) are specified using the `~~` operator:
